---
title: "lab2"
author: "Lydia Bleifuss"
date: "1/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(naniar)
library(VIM)
library(skimr)
library(ggfortify)
library(here)
```

## 1. Get some data

```{r}
ca_pb <- read_csv(here("data","ca_pollution_burden.csv"))

ca_dem <- read_csv(here("data","ca_census_demographics_2010.csv"))
```


## 2. Some data cleaning & wrangling 

```{r}
ca_pb_nopct <- ca_pb %>% 
  clean_names() %>% #makes everyting lower case and snake case 
  select(-contains("perc")) %>% #use minus to exclude things rather then select or keep only those, use ! if you want "do not match"
  select(-contains("pctl")) %>% 
  select(-latitude, -longitude)
```

```{r}
ca_dem_clean <- ca_dem %>% 
  clean_names()
```


## 3. Pollution burden indicators only 

```{r}
ca_pb_subset <- ca_pb_nopct %>% 
  select(ozone:solid_waste, asthma:housing_burden)
```


Now let's do PCA:
Use teh 'prcomp()' function
```{r, eval = FALSE}

#ERROR, good example 
pb_pca <- prcomp(ca_pb_subset, scale = TRUE) #error because we are missing data, needed to add eval = FALSE in the header so R doesn't try to run it when you knit 

```

Aside: exploring missingness!
```{r}
summary(ca_pb_subset) #you can use this to see where there are NA values

gg_miss_var(ca_pb_subset) #shows which variables have the most missingness, in this case it is "linguistic isolation"

matrixplot(ca_pb_subset) #higher values are darker and red indicates NAs 

matrixplot(ca_pb_subset, sortby = "poverty") #can also sort by a speciic variable, in this case "poverty" is organized from high to low  


```


Remove (complete case deletion) any observation with ANY NA values: #using "ANY NA, it will remove any ROW that has a missing variable"

```{r}
ca_pb_nona <- ca_pb_subset %>%
  drop_na()

summary(ca_pb_nona) #now we see that there aren't any NAs and there are appropriately less observations. Too much of a sacrifice to remove all of these values? Other remova NA options

skim(ca_pb_nona) #give a really cool summary that is better than summary funiton 

##### THIS SKIMER THING IS FREAKING AMAZING!!!!!###########
```


Now, no NA values, cool set to do PCA (otherwise, consider imputation, etc.): 

```{r}
my_ca_pca <- prcomp(ca_pb_nona, scale = TRUE) #normalizing the scales accross units 

my_ca_pca
summary(my_ca_pca) #shows how much (proportion) of the variance is explained by each principle component (PC#)

#biplot(my_ca_pca) #scary, moving on to autoplot

#going to use auto plot which is why we installed fortify 

my_biplot <- autoplot(my_ca_pca,
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE)+
scale_y_continuous(limits = c(-0.05, 0.05))

my_biplot

#this is only representing about 33% of varience, not great but...

#positively correlated: unemployment and poverty 
#no correlation: ozone and diesel_pm
#negatively correlation: drinking_water, groundwater_threats 

#because autoplot works with ggplot, you can use ggplot techniques to start customizing, but, lets look at what variables could be correlated to explor that 
```

## 4. PCA for polluiton burden + demographics 

Join together the ca_dem_clean with ca_pb_nopct and we will drop all of the NAs once we have joined them

```{r}
##you can always use full join to keep everything! Then you can decide what to drop. (MERGE is the SAME, it's just the base r version)

ca_df <- ca_dem_clean %>% 
  inner_join(ca_pb_nopct, by = c("census_tract_number" = "census_tract")) #use c because then you can add a comma and add more, the left of the = corresponds to the left dataframe

#Now drop NAs by complete case deletion:

ca_nona <- ca_df %>% 
  drop_na()

```

Make a subset of ca_nona that only includes: 

- white_percent
- elderly_65_percent
- pm2_5
- pesticides
- traffic
- asthma
- cardiovascular_disease
- poverty 

```{r}
my_sub <- ca_nona %>% 
  select(white_percent,white_percent, elderly_65_percent, pm2_5, pesticides, traffic, asthma, cardiovascular_disease, poverty)
```

PCA:
```{r}
my_dem_pca <- prcomp(my_sub, scale = TRUE)

my_dem_biplot <- autoplot(my_dem_pca,
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE)+
scale_y_continuous(limits = c(-0.05, 0.05))+
  theme_minimal()

my_dem_biplot
```

